Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-15 paper.tex
More references
Update section on cheminformatics toolkit. Add stubs for Indigo and RDKit. Remove comparison of toolkits as not relevant.

Open Source toolkits for cheminformatics have now existed for nearly
ten years. During this period, some toolkits were developed form
scratch, whereas others were made Open Source by releasing in-house
codebases under liberal licenses. The primary toolkits currently in
active development are the Chemistry Development Kit, Open Babel and
RDKit. These projects support a variety of languages natively (Java,
C++, Python), but by virtue of language wrappers, they can easily be
used from languages that they were not originally developed in. For
example, the CDK can be accessed from Python and Open Babel can be
accessed from Java via SWIG wrappers. One important aspect of the
toolkits noted here is that they originated for the support of desktop
and server-side applications.

More recently, a number of workers have developed light-weight
toolkits that are targeted towards web applications. One example of
this form is the ChemDoodle library from iChemLabs. The focus of these
toolkits is to support client-side cheminformatics functionality,
allowing for rich, chemically-aware browser-based applications.

In terms of functionality, most Open Source libraries support commonly
used cheminformatics features. At the same time, individual toolkits
have their own strengths. For example, the CDK has an extensive set of
molecular descriptors useful for QSAR modeling applications, but lacks
support for 3D structure generation. In contrast, Open Babel supports
3D structure generation and optimization using a variety of
forcefields, but lacks a built in set of molecular descriptors. RDKit
on the other hand supports both features and in general is the most
comprehensive Open Source toolkit in terms of functionality. Table XXX
summarizes the features of these three toolkits.

>>>>

\subsection*{Cheminformatics toolkits}

Open Source toolkits for cheminformatics have now existed for nearly
ten years. During this period, some toolkits were developed from
scratch in academia, whereas others were made Open Source by releasing in-house
codebases under liberal licenses. When the Blue Obelisk was
established five years ago, the primary toolkits under active development
were the Chemistry Development Kit (CDK)
\cite{Steinbeck2003, Steinbeck2006}, Open Babel \cite{WebOpenBabel},
and JOELib \cite{WebJOELib}. Of these, both the CDK and Open Babel
continue to be actively developed. [Insert main focus of work on CDK
in last 5 years] Since 2006, major new features of Open Babel include 3D structure
generation and 2D structure-diagram generation, UFF and MMFF94
forcefields, and significantly expanded support for computational
chemistry calculations. In addition, a major focus of Open Babel development
has been to fix problems in the areas of stereochemistry, kekulisation
and canonicalisation.

Two new Open Source cheminformatics toolkits have appeared since the
original paper. In 2006 Rational Discovery, a cheminformatics service
company (since closed down), released RDKit \cite{WebRDKit} under the
BSD License. RDKit continues to be actively developed and includes
code donated by Novartis. More recently, GGA Software Services (a
contract programming company) released the Indigo toolkit
\cite{WebIndigo} and associated software in 2009 under the GPL. [More
text on Indigo here]

One important aspect of the above
toolkits is that they originated for the support of desktop
and server-side applications. More recently, a number of workers have developed light-weight
toolkits that are targeted towards web applications. One example of
this form is the ChemDoodle library from iChemLabs. The focus of these
toolkits is to support client-side cheminformatics functionality,
allowing for rich, chemically-aware browser-based applications.
